---
title: "HRP 203: Module 3 Project"
author: "Valeria Gracia Olvera"
format: 
  pdf:
    fig-pos: 'H'
    geometry:
      - top=25mm
      - left=25mm
      - right=25mm
      - bottom=25mm
      - heightrounded
editor: 
  markdown: 
    wrap: sentence
bibliography: references.bib
fig-cap-location: bottom
---

```{r}
#| label: libraries
#| eval: true  # run code, default is TRUE
#| echo: false # output only
#| include: false
# Load libraries
library(ggplot2)
library(tidyverse)
library(GGally)
library(knitr)
library(tidyr)
library(jtools)
library(stargazer)
library(broom)
library(renv)
library(gtsummary)
```

[GitHub repository](https://github.com/vgraciaol/HRP203_repo)

# Introduction

Health care costs can be explained by factors such as age, gender, heart problems, and behaviors such as smoking.
Estimating the effect of these factors on healthcare costs could help allocate resources in the healthcare system.
In addition, it can also help estimate the economic impact of public health campaigns, such as smoking cessation campaigns.

The goal of this project is twofold: (1) estimate the effect of age, sex, heart problems, and smoking on health care costs; (2) evaluate the economic effect of smoking cessation campaigns.

# Methods

## Data

I used data provided in [Assignment 7](https://github.com/MethodsForReproducibleHealthResearch/Assignment7) which contains 5,000 simulated observations with `smoke`, `sex`, `age`, `cardiac`, and `cost` variables\footnote{I assumed the definition and description of the variables by looking at the data, but I have no more information to confirm it.} which are defined in @tbl-vars.

```{r}
#| label: tbl-vars
#| eval: true  # do not run code, default is TRUE
#| echo: false # output only

# Create data.frame
df_vars <- data.frame(variable = c("cost","age","female","cardiac","smoke"),
                      definition = c("Healthcare cost per year in USD", "Age in years", 
                                     "If =1 female","If =1 has heart problems","If =1 smoker"))

kable(df_vars, 
      col.names = c("Variable","Definition"),
      align = c("r","l"),
      caption = "Variable definition")
```

```{r}
#| label: load_data
#| eval: true  # run code, default is TRUE
#| echo: false # output only

# Load data from repository
df_cohort_raw <- read.csv(file="https://raw.githubusercontent.com/MethodsForReproducibleHealthResearch/Assignment7/main/raw-data/cohort.csv")

# Save raw data in repo
write.csv(df_cohort_raw, file = "data_raw/df_cohort_raw.csv")

# Copy of the data 
df_cohort <- df_cohort_raw %>% 
  rename(sex = female)

# Transform discrete to factors
df_cohort$smoke <- factor(df_cohort$smoke, levels = c(1,0), 
                          labels = c("Yes", "No"))
df_cohort$sex <- factor(df_cohort$sex, levels = c(1,0), 
                           labels = c("female", "male"))
df_cohort$cardiac <- factor(df_cohort$cardiac, levels = c(1,0), 
                            labels = c("Yes", "No"))
```

## Model

The relationship between the outcome and the explanatory variables can be written as: $$ cost_i = \beta_0 + \beta_1 age_i + \beta_2 I(female_i) + \beta_3 I(cardiac_i) + \beta_4 I(smoke_i) + \epsilon_i $$ where, $i$ indexes by individual, $I(\cdot)$ represents an indicator variable, and $\epsilon$ is the error term.

# Results

## Descriptive statistics

Descriptive statistics for the two continuous variables and the distribution between smokers or non-smokers (`smoke`), those who have heart problems (`cardiac`) or not, by sex, are presented in @tbl-descriptive-stats.
Note that the proportion of men and women who do not have heart problems is practically the same.
On the other hand, smokers and men are more likely to have heart problems and have higher healthcare costs.

```{r}
#| label: tbl-descriptive-stats
#| eval: true  # run code
#| echo: false # output only
#| message: false
#| tbl-cap: "Descriptive statistics by heart problems"

# Descriptive statistics
gtsummary::tbl_summary(df_cohort, by = c("cardiac"))
```

## Regression results

I estimated the effect of age, sex, heart problems, and smoking on the cost of healthcare.
@tbl-regression presents the regression results.

```{r}
#| label: tbl-regression
#| eval: true  # run code
#| echo: false # output only
#| message: false
#| tbl-cap: "Regression results"

# Regression
lm_cohort <- lm(cost ~ age + female + cardiac + smoke, data = df_cohort_raw)
gtsummary::tbl_regression(lm_cohort)
```

@fig-reg summarizes the relationship between the cost and the other variables of interest.
We can see that the cost increases with age for men who smoke and have heart problems.

```{r fig.height = 6, fig.width = 6}
#| label: fig-reg
#| eval: true  # run code
#| echo: false # output only
#| fig-cap: "Relationship between cost and all the other variables"
#| warning: false
 
df_plot <- df_cohort_raw %>% 
  rename(sex = female)

df_plot$smoke <- factor(df_plot$smoke, levels = c(1,0), 
                        labels = c("Smoke: yes", "Smoke: no"))
df_plot$sex <- factor(df_plot$sex, levels = c(1,0), 
                      labels = c("Female", "Male"))
df_plot$cardiac <- factor(df_plot$cardiac, levels = c(1,0), 
                          labels = c("Heart problems: yes", "Heart problems: no"))

ggplot(df_plot, aes(x = age, y = cost, color = cardiac)) +
  geom_point(size = 2, alpha = 0.10) +
  scale_colour_manual("", values = c("gray30", "steelblue")) +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(smoke ~ sex) +
  geom_smooth(method = "lm", formula = y ~ x, se = F,
              linewidth = 1.5) +
  xlab("Age (years)") + ylab("Cost ($)") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 0.5, 
                                   vjust = 0.5,
                                   colour = "black"),
        axis.text.y = element_text(colour = "black"),
        panel.border = element_rect(colour = "black", 
                                    fill = NA, 
                                    size = 0.7), 
        strip.background = element_rect(fill   = "transparent",
                                        colour = "transparent"),
        legend.justification = "center", 
        legend.position = "bottom",
        legend.text = element_text(size = 12),
        legend.direction = "horizontal", 
        legend.key = element_rect(fill   = "transparent", 
                                  colour = "transparent",
                                  size   = unit(3, "cm")))
```

## Results of a campaign to stop smoking

Finally, I am going to estimate the economic impact of a smoking cessation campaign under different scenarios:

```{=tex}
\begin{description}
  \item[Scenario A] The campaign is 60\% effective, i.e. 60\% of smokers quit smoking.
  \item[Scenario B] The campaign is 85\% effective, i.e. 85\% of smokers quit smoking.
\end{description}
```
I used the previously estimated model to predict the total cost in both scenarios and the status quo.
I randomly selected 60% of the smokers in the sample and changed their smoking status to non-smoking for scenario A.
I repeated this process now changing the smoking status of 85% of the smokers in the sample for scenario B.
Lastly, I calculated total healthcare costs by scenario and took the difference from the status quo to estimate the campaign's financial savings.

```{r}
#| label: analysis-cessation-campaign
#| eval: true  # run code
#| echo: false # output only

# Number of individuals in the sample
n_indiv <- dim(df_cohort_raw)[1]

# Create data.frame: status quo
df_cohort_SQ <- df_cohort_raw %>% 
  mutate(id = 1:n_indiv) %>% 
  relocate(id)

# Identify smokers and select randomly 60%/85% of them
id_smokers <- df_cohort_SQ$id[df_cohort_SQ$smoke == 1]
n_smokers <- length(id_smokers)
v_smokers_A <- sample(id_smokers, size = floor(0.60*n_smokers), replace = F)
v_smokers_B <- sample(id_smokers, size = floor(0.85*n_smokers), replace = F)

# Identify and change people who quit smoking -- create new data.frame
## Scenario A
df_cohort_scnA <- df_cohort_SQ %>% 
  mutate(quit_smoking = ifelse(id %in% v_smokers_A, 1, 0),
         smoke = ifelse(quit_smoking == 1, 0, smoke))

## Scenario B
df_cohort_scnB <- df_cohort_SQ %>% 
  mutate(quit_smoking = ifelse(id %in% v_smokers_B, 1, 0),
         smoke = ifelse(quit_smoking == 1, 0, smoke))

# Predict costs
## Scenario A
res_scnA <- predict(lm_cohort, newdata = df_cohort_scnA)
df_cohort_scnA$predict_cost <- res_scnA

## Scenario B
res_scnB <- predict(lm_cohort, newdata = df_cohort_scnB)
df_cohort_scnB$predict_cost <- res_scnB

## Status quo
res_scn <- predict(lm_cohort, newdata = df_cohort_SQ)
df_cohort_SQ$predict_cost <- res_scn

# Bind data.frames
df_predict_costs <- rbind(df_cohort_SQ %>% mutate(type = "Status quo"),
                          df_cohort_scnA %>% mutate(type = "Scenario A") %>% 
                            select(-quit_smoking),
                          df_cohort_scnB %>% mutate(type = "Scenario B") %>% 
                            select(-quit_smoking))

# Summarize
df_predict_costs_summ <- df_predict_costs %>% 
  group_by(type) %>% 
  summarise(total_cost = sum(predict_cost))

# Calculate difference from SQ
df_predict_costs_summ$diff_SQ <- df_predict_costs_summ$total_cost - df_predict_costs_summ$total_cost[df_predict_costs_summ$type == "Status quo"]

```

Results are presented in @fig-cessation-campaign.
Note that the highest savings comes from Scenario B, where `r round(-df_predict_costs_summ$diff_SQ[df_predict_costs_summ$type == "Scenario B"]/1000,0)` thousand USD are saved with the smoking campaign compared to status quo.

```{r fig.height=3, fig.width=3.5}
#| label: fig-cessation-campaign
#| eval: true  # run code
#| echo: false # output only
#| fig-cap: "Economic impact of a smoking cessation campaign, savings compared to status quo."

ggplot(subset(df_predict_costs_summ, type != "Status quo"),
       aes(x = type, y = -diff_SQ, color = type, fill = type)) +
  geom_bar(stat = "identity") +
  xlab("") + ylab("Savings ($)") +
  scale_colour_manual("", values = c("orange", "firebrick")) +
  scale_fill_manual("", values = c("orange", "firebrick")) +
  scale_y_continuous(labels = scales::comma) +
  labs(caption = "Scenario A: The campaign is 60% effective\n
       Scenario B: The campaign is 85% effective") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 0.5, 
                                   vjust = 0.5,
                                   colour = "black"),
        axis.text.y = element_text(colour = "black"),
        panel.border = element_rect(colour = "black", 
                                    fill = NA, 
                                    size = 0.7), 
        strip.background = element_rect(fill   = "transparent",
                                        colour = "transparent"),
        legend.justification = "center", 
        legend.position = "none",
        legend.text = element_text(size = 12),
        legend.direction = "horizontal", 
        legend.key = element_rect(fill   = "transparent", 
                                  colour = "transparent",
                                  size   = unit(3, "cm")))
```
